---
title: "Adjusting for response misclassification with the sslogit link"
subtitle: "Code and details for 'Bayesian models for missing and misclassified variables using integrated nested Laplace approximations'"
author: "Emma Skarstein, Leonardo Bastos, Håvard Rue and Stefanie Muff"
output: 
  pdf_document:
    extra_dependencies: ["bm"]
header-includes:
  - \usepackage{titling}
  - \pretitle{\begin{flushleft}\Huge\bfseries}
  - \posttitle{\end{flushleft}}  
  - \preauthor{\begin{flushleft}\Large}
  - \postauthor{\end{flushleft}}  
  - \predate{\begin{flushleft}\large}
  - \postdate{\end{flushleft}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
run_inla <- FALSE
```

# Accounting for imperfect sensitivity and specificity in the response

To use the `sslogit` link in INLA this needs to be enabled as follows:

```{r, message = FALSE}
library(INLA)
```

```{r}
inla.models <- INLA:::inla.get.inlaEnv()$inla.models
inla.models$link$sslogit$status <- NULL
assign("inla.models", inla.models, env = INLA:::inla.get.inlaEnv())
rm(inla.models)
```


Be aware that the reason this link is not available by default is that it struggles numerically if the sensitivity and specificity is large. SO USE WITH CAUTION.

```{r}
set.seed(1)
```


We simulate a misclassified response. `p_y` is the success probability for the correct version of the response, `y`. `s` is the misclassified response, and `p_s` is the success probability for `s`. 

```{r simulate data}
n <- 1000

p_y <- 0.1 # Success probability for y
sens <- 0.95 # P(y = 1 | s = 1)
spec <- 0.90 # P(y = 0 | s = 0)
p_s <- p_y*sens + (1-p_y)*(1-spec) # Success probability for s

df <- data.frame(s = rbinom(n = n, size = 1, prob = p_s))
```

Given this data, we can fit the naive model `r0`, which uses `s` without adjusting for misclassification, and the adjusted model `r1`, which adjusts for the misclassification by using and adjusted link function, as described in Section 4.1.

```{r fit M0 and M1, eval = run_inla}
formula <- s ~ 1

# Model 0 (ignoring sens and spec)
r0 <- inla(formula = formula, data = df, family = "binomial",
          Ntrials = 1)

# Model 1 (adding sens and spec)
r1 <- inla(formula = formula, data = df,
           family = "binomial",
           control.family = list(
             control.link = list(
               model = "sslogit",
               hyper = list(
                 sens = list(
                   prior = "logitbeta",
                   initial = inla.link.logit(sens),
                   fixed = TRUE
                 ),
                 spec = list(
                   prior = "logitbeta",
                   initial = inla.link.logit(spec),
                   fixed = TRUE)
               )
             )
           )
         )
```

```{r save M0 and M1, eval = run_inla, echo = FALSE}
saveRDS(r0, "results/r0.rds")
saveRDS(r1, "results/r1.rds")
```
```{r, load M0 and M1, echo = FALSE}
r0 <- readRDS("results/r0.rds")
r1 <- readRDS("results/r1.rds")
```

```{r evaluate M0 and M1}
r0$summary.fixed
r1$summary.fixed
```



# Alternative approach when sensitivity and specificity are not known exactly

For a third approach, we fit the model for a grid of different sensitivities and specificities, as described in the end of Section 4.1.

```{r create latin grid}
IC.sens = c(0.925, 0.975) 
IC.spec = c(0.85, 0.95)

# Latin square grid
n.latin <- 50
sens.values <- seq(from = IC.sens[1],
                   to = IC.sens[2],
                   length.out = n.latin)
spec.values <- seq(from = IC.spec[1], to = IC.spec[2],
                   length.out = n.latin)

# For the Latin-square
reord = sample(1:n.latin)
SSgrid = data.frame(sens = sens.values,
                    spec = spec.values[reord])
# Weights
# Using independent normals
sd.sens = (IC.sens[1] - sens) / qnorm(0.025)
sd.spec = (IC.spec[1] - spec) / qnorm(0.025)
SS.weights = dnorm(SSgrid$sens, sens, sd.sens) * dnorm(SSgrid$spec, spec, sd.spec)
SS.weights = SS.weights / sum(SS.weights)
```

We run INLA for each pair of sensitivity and specificity of the Latin square, and keep the results in a list:

```{r create M2, eval = run_inla}
inla.SSlogit.grid <- function(SS, data, formula){ 
  r.temp = inla(data = data,
                formula = formula, 
                family = "binomial", 
                control.family = list(
                  control.link = list( 
                    model = "sslogit", 
                    hyper = list(
                      sens = list(
                        prior = "logitbeta",
                        initial = inla.link.logit(SS$sens), 
                        fixed = TRUE),
                      spec = list(
                        prior = "logitbeta",
                        initial = inla.link.logit(SS$spec), 
                        fixed = TRUE)
                      )
                    )
                )
  )
  r.temp
}

r.list <- lapply(X = 1:nrow(SSgrid), 
                 FUN = function(x){
                   inla.SSlogit.grid(SSgrid[x,], 
                                     formula = formula,
                                     data = df)
                   } 
                 )

# Merging all INLA outputs
r2 <- inla.merge(loo = r.list, prob = SS.weights)
```
```{r save r2, echo = FALSE, eval=run_inla}
saveRDS(r2, "results/r2.rds")
```
```{r, echo = FALSE}
r2 <- readRDS("results/r2.rds")
```



# Comparing all approaches

The model that does no adjustment:
```{r}
r0$summary.fixed[,c(1,3,5)]
```

The model that adjusts for misclassification using fixed values for sensitivity and specificity:
```{r}
r1$summary.fixed[,c(1,3,5)]
```

The model that uses a grid of sensitivity and specificity:
```{r}
# Approximate 95% interval
c(r2$summary.fixed$mean, r2$summary.fixed$mean + c(-2, 2)*r2$summary.fixed$sd)

# Using inla.qmarginal
c(r2$summary.fixed$mean, 
  inla.qmarginal(p = c(0.025, 0.975), 
                 marginal = r2$marginals.fixed$`(Intercept)`))
```


```{r}
marginals.df <- dplyr::bind_rows(
  'M0' = data.frame(r0$marginals.fixed$`(Intercept)`), 
  'M1' = data.frame(r1$marginals.fixed$`(Intercept)`), 
  'M2' = data.frame(r2$marginals.fixed$`(Intercept)`), 
  .id = "Models")

marginals.df$Models <- factor(marginals.df$Models, levels = c("M0", "M1", "M2"))
marginals.df$Models <- plyr::revalue(marginals.df$Models, 
                               c("M0" = "Ignoring misclassification", 
                                 "M1" = "Fixed sens. and spec.", 
                                 "M2" = "Propagated uncertainty"))

library(ggplot2)

ggplot(marginals.df, aes(x = x, y = y, color = Models, fill = Models)) + 
  geom_area(alpha = 0.5, position = "identity") +
  geom_vline(xintercept = inla.link.logit(p_y), show.legend = F, 
             linetype = "dashed") +
  annotate("text", x = inla.link.logit(p_y) - 0.08, y = 3.8, label="True value", 
           hjust = 0, angle = 90) +
  scale_color_viridis_d() + scale_fill_viridis_d() +
  coord_cartesian(xlim = c(-3.3, -1)) +
  theme_bw() +
  xlab(bquote(p[y])) +
  theme(legend.position.inside = c(0.2, 0.8),
        axis.title.y = element_blank())
ggsave("figures/response_mc.pdf", height = 4, width = 7)
```



```{r, eval = FALSE, echo = FALSE}
# Code from Olivier Supplisson
library(INLA)
inla.models <- get("inla.models", env = INLA:::inla.get.inlaEnv())
inla.models$link$sslogit$status <- NULL
assign("inla.models", inla.models, env = INLA:::inla.get.inlaEnv())
rm(inla.models)

n = 10000
size = 1
x = rnorm(n, s=0.2)
eta = 2 + x
sens = 0.85
spec = 0.995
fixed = TRUE
p = 1/(1+exp(-eta))
prob = sens * p + (1-spec)*(1 - p)
y = rbinom(n, size=size,  prob=prob)

a = 100
b = function(s, a) a*(1-s)/s

r.naive = inla(y ~ 1 + x, data = data.frame(y, x, size),
         family = "binomial",
         Ntrials = size)

r = inla(y ~ 1 + x, data = data.frame(y, x, size),
         family = "binomial",
         Ntrials = size,
         verbose = TRUE,
         inla.mode = "experimental",
         control.inla = list(cmin = 1, b.strategy = "keep"),
         control.family = list(
           control.link = list(
             model = "sslogit",
             hyper = list(
               sens = list(
                 prior = "logitbeta",
                 initial = inla.link.logit(sens),
                 fixed = fixed,
                 param = c(a, b(sens, a))),
               spec = list(
                 prior = "logitbeta",
                 initial = inla.link.logit(spec),
                 fixed = fixed,
                 param = c(a, b(spec, a)))))))

r$.args$control.inla$cmin <- 0
r <- inla.rerun(r)
summary(r)

r.naive$summary.fixed
r$summary.fixed

```

